---
title: "Project"
author: "Group 33"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(ggbiplot)
```

# Preparing the data
```{r}
columnNames <- c("id","diagnosis","radius_mean","texture_mean","perimeter_mean",
                 "area_mean","smoothness_mean","compactness_mean","concavity_mean",
                 "concave_points_mean","symmetry_mean","fractal_dimension_mean",
                 "radius_se","texture_se","perimeter_se","area_se","smoothness_se",
                 "compactness_se","concavity_se","concave_points_se","symmetry_se",
                 "fractal_dimension_se","radius_worst","texture_worst","perimeter_worst",
                 "area_worst","smoothness_worst","compactness_worst","concavity_worst",
                 "concave_points_worst","symmetry_worst","fractal_dimension_worst")
```

```{r}
wisc.df <- fread("wdbc.data", col.names = columnNames)
head(wisc.df)
```
# Check for missing data

```{r}
# Sum of NA values in each column
colSums(is.na(wisc.df))
```
# Exploratory Data Analysis

Our response variable is diagnosis: Benign (B) or Malignant (M). We have 3 sets of 10 numeric variables: mean, se, worst Let’s first collect all the 30 numeric variables into a matrix.


```{r}
# Convert the features of the data: wisc.data
wisc.data <- as.matrix(wisc.df[, .SD, .SDcols = 3:32])  # Selects only the feature columns
```
```{r}
# Set the row names of wisc.data
row.names(wisc.data) <- wisc.df$id
```
```{r}
# Create diagnosis vector
diagnosis <- as.numeric(wisc.df$diagnosis == "M")
```

# Number of observations in this dataset.

```{r}
nrow(wisc.data)
```
# Number of features in the data are suffixed with _mean, _se, _worst?

```{r}
sum(endsWith(colnames(wisc.data), "_mean"))
sum(endsWith(colnames(wisc.data), "_se"))
sum(endsWith(colnames(wisc.data), "_worst"))
```

# Number of observations have benign or malignant diagnosis.

```{r}
library(ggplot2)
# Create a frequency table of the diagnosis
diagnosis_count <- table(wisc.df$diagnosis)
diagnosis_count

```
```{r}
# Convert the table to a data frame for ggplot2
diagnosis_df <- as.data.frame(diagnosis_count)
names(diagnosis_df) <- c("Diagnosis", "Count")

# Create a bar plot using ggplot2
ggplot(diagnosis_df, aes(x = Diagnosis, y = Count, fill = Diagnosis)) +
  geom_bar(stat = "identity") +  # Use identity to use the actual counts
  theme_minimal() +  # Use a minimal theme for a cleaner look
  labs(x = "Diagnosis Type", y = "Frequency", title = "Distribution of Diagnosis in Dataset") +
  scale_fill_brewer(palette = "Set1")  # Color palette for visual distinction

```

# Correlation

```{r}
corMatrix <- wisc.df[,c(3:32)]

# Rename the colnames
cNames <- c("rad_m","txt_m","per_m",
                 "are_m","smt_m","cmp_m","con_m",
                 "ccp_m","sym_m","frd_m",
                 "rad_se","txt_se","per_se","are_se","smt_se",
                 "cmp_se","con_se","ccp_se","sym_se",
                 "frd_se","rad_w","txt_w","per_w",
                 "are_w","smt_w","cmp_w","con_w",
                 "ccp_w","sym_w","frd_w")

colnames(corMatrix) <- cNames

# Create the correlation matrix
M <- round(cor(corMatrix), 2)

# Create corrplot
library(corrplot)
corrplot(M, diag = FALSE, method="color", order="FPC", tl.srt = 90)

```

# Performing PCA

# Check column means and standard deviations

```{r}
colMeans(wisc.data)
```

```{r}
apply(wisc.data, 2, sd)
```


```{r}
# Execute PCA, scaling if appropriate: wisc.pr
wisc.pr <- prcomp(x = wisc.data, scale = TRUE)

# Look at summary of results
summary(wisc.pr)
```
```{r}
screeplot(wisc.pr, type = "line", main = "Scree Plot for PCA")
```

# Interpreting PCA results

```{r}
# Create a biplot of wisc.pr
biplot(wisc.pr, scale = 0)
```



```{r}
#Scatter plot observations by components 1 and 2
plot(wisc.pr$x[, c(1, 2)], col = (diagnosis+1), 
     xlab = "PC1", ylab = "PC2")
```


```{r}
# Repeat for components 1 and 3
plot(wisc.pr$x[, c(1, 3)], col = (diagnosis + 1), 
     xlab = "PC1", ylab = "PC3")
```


```{r}
plot(wisc.pr$x[, c(1, 4)], col = (diagnosis + 1), 
     xlab = "PC1", ylab = "PC4")
```
# Variance explained

```{r}

# Calculate variability of each component
pr.var <- wisc.pr$sdev^2

# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "b")
```
```{r}

# Plot cumulative proportion of variance explained
plot(cumsum(pve), xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "b")
```
```{r}
# Extract principal components
PC1 <- wisc.pr$x[, 1]
PC2 <- wisc.pr$x[, 2]

# Create the biplot
biplot <- ggbiplot(wisc.pr, 
                   obs.scale = 1, 
                   var.scale = 1) +
          theme_minimal()

# Add point layer with colors based on smoking status
biplot + geom_point(aes(x = PC1, y = PC2, color = factor(diagnosis)), size = 2)
```
```{r}
# Convert PCA results to a data frame
pca_df <- as.data.frame(wisc.pr$x)
```



# Clustering

# Hierarchical clustering 

```{r}
# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)


# Calculate the (Euclidean) distances: data.dist
data.dist <- dist(data.scaled)

# Create a hierarchical clustering model: wisc.hclust
wisc.hclust = hclust(data.dist, method = "complete")
```

```{r}
library(ggdendro)
ggdendrogram(wisc.hclust, segements=TRUE, labels=TRUE, leaf_labels = TRUE, rotate=FALSE, theme_dendro = TRUE) +
 labs(title='Complete Linkage')
```


```{r}
# Cut tree so that it has 4 clusters: wisc.hclust.clusters
wisc.hclust.clusters <- cutree(wisc.hclust, k = 4)

# Compare cluster membership to actual diagnoses
table(wisc.hclust.clusters, diagnosis)
```
```{r}
# Create a matrix to represent the updated counts from the clustering results
# Rows represent clusters, columns represent actual diagnosis
updated_matrix <- matrix(c(12, 165, 2, 5, 343, 40, 0, 2), 
                         nrow = 4, byrow = TRUE,
                         dimnames = list(c("Cluster 1", "Cluster 2", "Cluster 3", "Cluster 4"),
                                         c("Benign", "Malignant")))

# Calculate the total number of correct predictions
# Assuming Cluster 1, 2, and 4 are meant to predict malignant, and Cluster 3 predicts benign
correct_predictions <- updated_matrix[1, 2] +  # Malignant correctly predicted in Cluster 1
                       updated_matrix[2, 2] +  # Malignant correctly predicted in Cluster 2
                       updated_matrix[3, 1] +  # Benign correctly predicted in Cluster 3
                       updated_matrix[4, 2]    # Malignant correctly predicted in Cluster 4

# Calculate total predictions
total_predictions <- sum(updated_matrix)

# Calculate accuracy
accuracy <- correct_predictions / total_predictions

# Print the accuracy
print(accuracy)

```

# K-Means Clustering

```{r}
set.seed(143)
km_out_list <- lapply(1:10, function(k) list(
  k=k,
  km_out=kmeans(wisc.data, k, nstart = 20)))

km_results <- data.frame(
  k=sapply(km_out_list, function(k) k$k),
  totss=sapply(km_out_list, function(k) k$km_out$totss),
  tot_withinss=sapply(km_out_list, function(k) k$km_out$tot.withinss)
  )
km_results

ggplot(km_results,aes(x=k,y=tot_withinss))+geom_line()+geom_point()
```
```{r}
set.seed(143)
library(factoextra)
fviz_nbclust(wisc.data, kmeans, method = "wss",k.max=10, nstart=20, iter.max=20) +
  geom_vline(xintercept = 2
             , linetype = 2)+
  labs(subtitle = "Elbow method")
```
```{r}
set.seed(143)
library(cluster)
library(plotly)

# Perform k-Means clustering
km_out <- kmeans(data.scaled, centers = 2, nstart = 25)

plot_ly(x=wisc.pr$x[,1],y=wisc.pr$x[,2], color = as.factor(km_out$cluster), colors=c("red","blue"))
```



```{r}
# Create a k-means model on wisc.data: wisc.km
wisc.km <- kmeans(scale(wisc.data), centers = 2, nstart = 20)

# Compare k-means to actual diagnoses
table(wisc.km$cluster, diagnosis)
```
```{r}
# Create a matrix or data table to represent the counts from the clustering results
# Diagnosis (0 for benign, 1 for malignant)
# Rows represent clusters, columns represent actual diagnosis
results_matrix <- matrix(c(14, 175, 343, 37), nrow = 2, byrow = TRUE)

# Assign row names and column names for clarity
rownames(results_matrix) <- c("Cluster 1", "Cluster 2")
colnames(results_matrix) <- c("Benign", "Malignant")

# Calculate the total number of correct predictions
# Correct predictions for Cluster 1 (Malignant) and Cluster 2 (Benign)
correct_predictions <- results_matrix[1, 2] + results_matrix[2, 1]

# Calculate total predictions
total_predictions <- sum(results_matrix)

# Calculate accuracy
accuracy <- correct_predictions / total_predictions

# Print the accuracy
print(accuracy)

```


# Clustering on PCA results


```{r}
# Create a hierarchical clustering model: wisc.pr.hclust
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "complete")
```


```{r}
# Cut model into 4 clusters: wisc.pr.hclust.clusters
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 4)
# Compare to actual diagnoses
table( wisc.pr.hclust.clusters, diagnosis)
```
```{r}
# Create a matrix to represent the updated counts from the clustering results
# Rows represent clusters, columns represent actual diagnosis
updated_matrix <- matrix(c(5, 113, 350, 97, 2, 0, 0, 2), 
                         nrow = 4, byrow = TRUE,
                         dimnames = list(c("Cluster 1", "Cluster 2", "Cluster 3", "Cluster 4"),
                                         c("Benign", "Malignant")))

# Calculate the total number of correct predictions
# Assuming Cluster 1 and 4 are meant to predict malignant, and Cluster 2 and 3 predict benign
correct_predictions <- updated_matrix[1, 2] +  # Malignant correctly predicted in Cluster 1
                       updated_matrix[2, 1] +  # Benign correctly predicted in Cluster 2
                       updated_matrix[3, 1] +  # Benign correctly predicted in Cluster 3
                       updated_matrix[4, 2]    # Malignant correctly predicted in Cluster 4

# Calculate total predictions
total_predictions <- sum(updated_matrix)

# Calculate accuracy
accuracy <- correct_predictions / total_predictions
#accuracy= 0.9321763
# Print the accuracy
print(accuracy)
```




